Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-04-06 paper.qmd
removed old orbital figures; regenerate data cubes figure; handle SI units properly

format:
  ieee-pdf:
    keep-tex: true
    journaltype: conference
    papersize: a4
  ieee-html: default

@@

We base our hourly flight density estimation on (enhanced) trajectories flown
by \acr{IFR} commercial aircraft as
recorded by EUROCONTROL's \acr{NM}[^1] and overlapped on Uber's H3
gridding system at different resolutions [@h3resolutions], see @fig-cells 
for resolution 2 which corresponds to hexagons of average area $86801.78\ km^2$
and average edge length of $182.51 km$, see also table

@@

```{r cells-map}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/cells_2.png")
```

@@

Flight density in the EUROCONTROL Area on 2024-08-01T12 (H3 cells at resolution 3).
:::

@@

@@

From Equation 15 in [@patera2008], the (on the ground) impact probability
density, $\sigma(L, i)$, at latitude $L$ of a reentring object
whose orbit has inclination $i$ (see @fig-orbital-inclination) is given by:

$$ \sigma(L, i) = \frac{1}{2 \pi^2 R^2 \sqrt{(\sin^2(i) - \sin^2(L))}}$$

where $R$ is the Earth's radius.

::: {#fig-orbital-inclination}

```{r orbital-inclination}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/orbital_inclination_clip.png")
```

Orbital inclination is the angle between the plane of an orbit and the equator. An orbital inclination of 0° is directly above the equator, 90° crosses right above the pole, and 180° orbits above the equator in the opposite direction of Earth’s spin. (NASA illustration by Robert Simmon.).
:::

@@

$$
\varepsilon_i = \frac{s_{a} a_{h} + s_{d} a_{v}}{s_d}
$$
where $s_{a}$ is the average aircraft speed, $s_{d}$ is the average debris 
object fall speed, $a_v$ area of the aircraft as viewed from above,
$a_h$ area of the aircraft as viewed from the from the front of the aircraft.

@@

```{r densities}
#| out-width: "120%"
#| echo: false
knitr::include_graphics("media/figures/cubes.png")
```

Flight and population densities data (hex on Germany as an example).
:::

@@

# References

>>>>

format:
  ieee-pdf:
    keep-tex: true
    # fig-pos: 'H'
    journaltype: conference
    papersize: a4
    include-in-header:
      - file: media/preamble.tex
    include-before-body:
      - text: |
          % NOTE: make siunitx's \qty work without clashing with the one in physics
          \let\qty\SI
  ieee-html: 
    html-math-method: mathjax
    include-in-header:
      file: media/siunitx.html

@@

We base our hourly flight density estimation on trajectories flown
by \acr{IFR} commercial aircraft as
recorded by EUROCONTROL's \acr{NM}[^1] and overlapped on Uber's H3
gridding system at different resolutions [@h3resolutions].
@fig-cells shows cells for resolution $2$ which corresponds to hexagons
of average area $\qty{86801.78}{km^2}$ and average edge length of 
$\qty{182.51}{km}$, see also @tbl-resolutions for other resolutions.

@@

```{r cells-map}
#| out-width: "100%"
#| echo: false
#| fig-scap: "H3 cells at resolution 2"
knitr::include_graphics("media/figures/cells_2.png")
```

@@

Flight density in the EUROCONTROL Area on 2024-08-01T12 (H3 cells at resolution $3$).
:::

@@

The probability of impact for debris object $k$, $p_k$, is expressed as.

$$
p_k = \varepsilon_k \int_a \rho(\vec{x}) \sigma(\vec{x}) da
$$ {#eq-impact-probability}

where $\rho({\vec{x}})$ is the probability density of an aircraft
and $\sigma(\vec{x})$ is probability density of a debris object in the
position $\vec{x}$.

@@

Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
circular orbits prior to reentry even if initially on an elliptical ones.
In fact drag at perigee will lower apogee and similarly drag at apogee will
lower perigee. But because drag at perigee is higher due to denser atmosphere,
the apogee  will decrease faster than perigee and the orbit will then tend to be
a circular one.
The angular position of the reentry point from the ascending node, 
see @sec-orbitals, is uniformly distributed, $\sigma(\theta) = 1 / (2\pi)$,
due to uncertainties in atmospheric drag and the gravitational effects of
the Equatorial bulge, hence we assume that the impact longitude, $\lambda$,
is uniformely distributed $\sigma(\lambda) = 1/(2\pi)$.

The impact probability density, $\sigma(\varphi, i)$,
at latitude $\varphi$ of a reentring object whose orbit has inclination $i$
(see @fig-orbital-elements) is given by [[Equation 15 in @patera2008]]:

$$ \sigma(\varphi, i) = \frac{1}{2 \pi^2 R^2 \sqrt{(\sin^2(i) - \sin^2(\varphi))}}$$ {#eq-impact-probability-density}

So for example [@patera2008], if a building is located at latitude 30 degrees
north (or south) and has an area of $\qty{1000}{m^2}$ and a piece of debris
reenters the Earth’s atmosphere from an orbit having an inclination of 45 degrees,
the impact probability is given by

$$\begin{aligned}
P &= \Delta A \times \sigma(\varphi, i) = 1000 \times \sigma\Bigl(\frac{\pi}{6}, \frac{\pi}{4}\Bigl) \\
  &= 1000 \times \num{2.490756e-15} = \num{2.490756e-12}
\end{aligned}$$

where we used the equatorial value of $\qty{6378}{km}$, for the Earth's radius,
$R$, as the distance of the building from the center of the planet.

$\num{0.046}$

@@

The effective expose area, $\varepsilon_i$, is
$$
\varepsilon_i = \frac{s_{a} a_{h} + s_{d} a_{v}}{s_d}
$$
where $s_{a}$ is the average aircraft speed, $s_{d}$ is the average debris 
object fall speed, $a_v$ area of the aircraft as viewed from above,
$a_h$ area of the aircraft as viewed from the from the front of the aircraft.

@@

```{r densities}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/cubes.png")
```

Data cube of hourly flight densities data over H3 hexes (in Germany as an example) for multiple dates, $d_1 \cdots d_n$. 
:::

@@

# Appendix {.appendix}

## Orbital Elements {#sec-orbitals .appendix}
We refer to @fig-orbital-elements to describe the orbital parameters of
interest for the current work.

:::{#fig-orbital-elements layout="[40,-3,57]" fig-env="figure*"}

![Keplerian elements: semi-major axis, $a$, the eccentricity, $e = \frac{2c}{2a}$. (the semiparameter (a.k.a. _semilatus rectum_), $p$, describes the size of the conic section by defining the width at the primary focus, and is defined as
$p = b^2 / a = a (1 - e^2)$)](media/figures/classical_orbital_elements_2.png)

![Keplerian elements: inclination, $i$, the right ascension of the ascending node, $\Omega$ , and the argument of periapsis, $\omega$.(Earth rotates around $\hat{K}$)](media/figures/classical_orbital_elements_4.png)

Definition of Keplerian elements. The orbital elements describing the trajectory of orbiting object are illustrated in association with an inertial reference frame $\hat{I}$-$\hat{J}$-$\hat{K}$. The $I$-axis, $\hat{I}$, points toward the vernal equinox and the $J$-axis, $\hat{J}$, is perpendicular to the $I$-axis in the equatorial plane n the sense of Earth's rotation with the $K$-axis pointing up. The semi-major axis, $a$, the eccentricity, $e$, the inclination, $i$, the right ascension of the ascending node, $\Omega$ , and the argument of perigee, $\omega$, constitute, together with the mean anomaly at the initial position, $M_0$, a set of six constants of motion that are preserved in the absence of perturbation. True anomaly, $\nu$, specifies the location of an orbiting object within the orbit. It is the angle between perigee and the orbiting object’s position vector measured in the direction of the orbiting object’s motion.
Figures from [@sellers2003understanding], respectively Figure 5-2 and Figure 5-9.

:::

The angular position, $\theta$, of the reentry point from the ascending node
in the orbit plane is $\theta = \omega + \nu$.
It is related to the latitude, $\varphi$, via the Cartesian coordinate $k$,
$k = R \sin(\varphi) = R \sin(\theta)\sin(i)$, or

$$\sin(\varphi) = \sin(\theta)\sin(i)$$ {#eq-spheric-coords}

Mass of Earth, $M_{\oplus} = \qty{5.9722e24}{kg}$.

# References {.unnumbered}
